---
title: "ggplot2"
author: "Anthony Hung"
date: "2020-02-20"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

# Data visualization in many dimensions with ggplot2

# Load data and libraries

```{r}
library(ggplot2)
cars <- mtcars
?mtcars
```

# Examine structure of data

These data contain 32 observations on cars across 11 different variables. Some of these variables (e.g. mpg) are numeric, while others (e.g. cyl) are factors. Notice that these data are in "tidy" format, meaning that:

1. Each variable forms a column.

2. Each observation forms a row.

3. Each type of observational unit forms a table.

```{r}
dim(cars)
head(cars)
```

# Plot showing 2 dimensions

Let's examine the relationship between two variables, mpg and wt: 

```{r}
ggplot(data = cars) + #data layer
  geom_point(aes(x = wt, y = mpg)) #geom layer, aesthetics layer
```

# Plot showing 3 dimensions

Now let's color by cylinder number. What is important to take into account for the cylinder variable?

```{r}
ggplot(data = cars) + #data layer
  geom_point(aes(x = wt, y = mpg, color = cyl)) #geom layer, aesthetics layer

ggplot(data = cars) + #data layer
  geom_point(aes(x = wt, y = mpg, size = factor(cyl))) #geom layer, aesthetics layer, color
ggplot(data = cars) + #data layer
  geom_point(aes(x = wt, y = mpg, shape = factor(cyl))) #geom layer, aesthetics layer, color
ggplot(data = cars) + #data layer
  geom_point(aes(x = wt, y = mpg, color = factor(cyl))) #geom layer, aesthetics layer, color
```

# Plot showing 4 dimensions

What if we're interested in the same plot as above, but separated by automatic vs manual transmissions?

```{r}
ggplot(data = cars) + #data layer
  geom_point(aes(x = wt, y = mpg, color = factor(cyl))) + #geom layer, aesthetics layer, color
  facet_wrap(~factor(am)) #facet

#rename factor labels
cars$am <- as.factor(cars$am)
levels(cars$am) <- c("automatic", "manual")

ggplot(data = cars) + #data layer
  geom_point(aes(x = wt, y = mpg, color = factor(cyl))) + #geom layer, aesthetics layer, color
  facet_wrap(~am) #facet
  
```

# Plot showing 5 dimensions

```{r}
cars$vs <- as.factor(cars$vs)
levels(cars$vs) <- c("V-shaped engine", "straight engine")

ggplot(data = cars) + #data layer
  geom_point(aes(x = wt, y = mpg, color = factor(cyl))) + #geom layer, aesthetics layer, color
  facet_grid(vs~am) #facet
```

# Gone too far?

```{r}
ggplot(data = cars) + #data layer
  geom_point(aes(x = wt, y = mpg, color = factor(cyl), size = disp, shape = factor(gear))) + #geom layer, aesthetics layer, color
  facet_grid(vs~am)
```

# Additional points

## Adding labels

```{r}
labeled_plot <- ggplot(data = cars) + #data layer
  geom_point(aes(x = wt, y = mpg, color = factor(cyl))) + #geom layer, aesthetics layer, color
  ggtitle("Relationship between mpg and weight \n in mtcars") + 
  xlab("weight (1000 lbs)") + 
  ylab("miles per gallon (mpg)") +
  labs(color = "Cylinder type")

labeled_plot
```


## Choosing colors

```{r}
cylinder_colors <- c("#035AA6", "#F2AE2E", "#F23D3D")

labeled_plot +
  scale_color_manual(values=cylinder_colors)
```

## Reordering factors

```{r}
cars$cyl <- factor(cars$cyl)
cars$cyl <- factor(cars$cyl, levels(cars$cyl)[c(3,2,1)])

ggplot(data = cars) + #data layer
  geom_point(aes(x = wt, y = mpg, color = cyl)) + #geom layer, aesthetics layer, color
  ggtitle("Relationship between mpg and weight \n in mtcars") + 
  xlab("weight (1000 lbs)") + 
  ylab("miles per gallon (mpg)") +
  labs(color = "Cylinder type") +
  scale_color_manual(values=cylinder_colors)
```

## Adding statistics

Remember, we are working with the grammar of graphics so we can add as many geoms as we want/need to our plot!

```{r}
labeled_plot + 
  geom_smooth(aes(x = wt, y = mpg, color = factor(cyl)), method = "lm", se = FALSE) #additional geoms inherit data and aesthetics from the predefined plot


#we can also define a new dataset for an additional geom
library(dplyr)
means <- cars %>% 
  group_by(cyl) %>% 
  summarise(mean.wt = mean(wt), mean.mpg = mean(mpg))
means

labeled_plot + 
  geom_point(data = means, aes(x = mean.wt, y = mean.mpg, color = cyl), size = 10, alpha = 0.5) +
  scale_color_manual(values=c(cylinder_colors))
```

## Themes

```{r}
library(ggthemes)
labeled_plot + 
  theme_fivethirtyeight()
```

