---
title: "Introduction to R"
author: "Anthony Hung"
date: "2019-04-25"
output: 
  workflowr::wflow_html:
    code_folding: show
editor_options:
  chunk_output_type: console
---


<style>
hide {
  background-color: #d6d6d6;
  color: #d6d6d6;
}
hide:hover {
  background-color: white;
  color: black;
}
</style>

## Introduction

Here, we introduce R, a statistical programming language. Doing statistics within a programming language brings many advantages, including allowing one to organize all analyses into program files that can be rerun to replicate analyses. In addition to using R, we will be using RStudio, an integrated development environment (IDE), which assists us in working with R and outputs of our code as we develop it. Our objective today is to get everyone up to speed with working knowledge of R and programming to be able to do exercises as a part of the rest of the course.

Both R and RStudio are freely available online.

## Downloading/Installing R and RStudio

* Download the appropriate "base" version of R for your operating system from CRAN: https://cran.r-project.org/

* Install the software with default settings.

* Download the appropriate RStudio version for your operating system: https://www.rstudio.com/products/rstudio/download/#download

## R Basics

Follow along in your R console with the code in each of the code chunks as we explore the different aspects of R! Clicking on the github logo on the top right corner of the webpage will take you to the repository for this website, where you can download the R markdown file for this page to load into RStudio to follow along.

### Mathematical operations in R

Many familiar operators work in R, allowing you to work with numbers like you would in a calculator. Operators such as inequalities also work, returning "TRUE" if the proposed logical expression is true and "FALSE" otherwise.

```{r math}
2+4 #addition
2-4 #subtraction
2*4 #multiplication
2/4 #division
2^4 #exponentiation
log(2) #the default log base is the natural log

2 < 4
2 > 4
2 >= 4 #greater than or equal to 
2 == 2 #is equal to (notice that there are two equal signs, as a single equal sign denotes assignment)
2 != 4 #is not equal to 
2 != 4 | 2 + 2 == 4 #OR
2 != 4 & 2 + 2 == 4 #AND
"Red" == "Red"
```

### Objects

In addition to being able to work with actual numbers, R works in objects, which can represent anything from numbers to strings to vectors to matrices. Everything in R is an object. The best practice for assigning variable names to objects is the "<-" operator. After objects are created, they are stored in in the "Environment" tab in your RStudio console and can be called upon to perform different operations.

R has many data structures, including:

* atomic vector
* list
* matrix
* data frame
* factors

R has 6 atomic vector types, or classes. Atomic means that a vector only contains elements of one class (i.e. the elements inside the vector do not come from mutliple classes).

* character
* numeric (real or decimal)
* integer
* logical (TRUE or FALSE)
* complex (containing i)

```{r objects}
a <- 2
b <- 3
a + b
class(a) #the "class" function tells you what class of object a is

d <- c(1,2,3,4,5) #the "c" function concatenates the arguments contained within it into a vector
d
d <- c(d, 1) #The "c" function also allows you to append items to an existing vector
d
class(d)
d[3] #brackets allow you index vectors or matrices. Here, we call the third value from our d vector.

#Matrices are just like vectors, but with two dimensions
my_matrix <- matrix(seq(1:9), ncol = 3)
my_matrix

#vectors and matrices can only contain objects of one class. If you include objects of multiple types into the same vector, R will perform coersion to force all the objects contained in the vector into a shared class
x <- c(1.7, "a")
x
class("a")
class(1.7)
class(x)
y <- c(TRUE, 2)
y
z <- c("a", TRUE)
z

#If you would like to store objects of multiple classes into one object, a list can accomodate such a task.
x_y_z_list <- list(x,y,z)
x_y_z_list
#to index an element in a list, use double brackets [[]]. You can further index elements within an element of a list.
x_y_z_list[[1]]
x_y_z_list[[1]][2]
#elements in a list can be assigned names
x_y_z_list <- list(a=x, b=y, c=z)
x_y_z_list

#Dataframes are a very commonly used type of object in R. You can think of a dataframe as a rectangular combination of lists.
#The below code stores the stated values in a dataframe which contains employee ids, names, salaries, and start dates for 5 employees
emp.data <- data.frame(
   emp_id = c (1:5), 
   emp_name = c("Rick","Dan","Michelle","Ryan","Gary"),
   salary = c(623.3,515.2,611.0,729.0,843.25), 
   
   start_date = as.Date(c("2012-01-01", "2013-09-23", "2014-11-15", "2014-05-11",
      "2015-03-27")),
   stringsAsFactors = FALSE
)

emp.data
emp.data$emp_id #the $ operator calls on a certain column of a dataframe
class(emp.data$emp_id) #As noted earlier, a dataframe can be thought of as a rectangular list, combining different data classes together, each in a different column.
class(emp.data$salary)
emp.data$emp_name[emp.data$salary > 620] #You can combine logical operators, brackets, and the $ sign to subset your dataframe in any way you choose! Here, we print out all the employee names for employees who have a salary greater than 620.

ls() #ls lists all the variable names that have been assigned to objects in your workspace
```

### Using Packages in R

In addition to the basic functions provided in R, oftentimes we will be working with packages that contain functions written by other people to perform common tasks or specific analyses. Packages can also contain datasets. We can load these packages into our R environment after installing them in R.

```{r Packages}
# note that we won't use most of these packages today, but it's nice to have them for later on in the course
install.packages("dplyr")
install.packages("ggplot2")
install.packages("cowplot")
install.packages("grid")
install.packages("e1071")
install.packages("caret")
install.packages("gapminder")
install.packages("pROC")

library("gapminder") #After installing the package, we need to tell R to load it into our current environment with this function.
head(gapminder) #The package gapminder contains a dataset called gapminder. We can use the "head" function to print out the first 6 rows of this dataset.

?gapminder #the ? operator lauches a help page to describe a particular function, including the arguments it takes. Whenever using a new function, it is good practice to first explore it through ?.
```

### Loops

Oftentimes, we may want to perform the same operation or function many many times. Rather than having to explicitly write out each individual operation, we can make use of loops. For example, let's say that we want to raise the number 2 to the power of each integer from 0 to 20. We could either write out 2^0, 2^1, 2^2 ..., or make use of a for loop to condense our code while getting the same result.

```{r loops}
2^0
2^1
2^2
2^3
# ...

#This is a for loop. in the parentheses after the for function, we specify over what range of values we want to loop over, and assign a dummy variable name to take on each of those values in sequence. Within the curly braces, we state what operation we want to perform over all the values taken on by the dummy variable.
for(i in 0:20){
  print(2^i)
}
```

### User-defined Functions (UDF)

Another way to avoid writing out or copy-pasting the same exact thing over and over again when working with data is to write a function to contain a certain combination of operations you find yourself running mutliple times. For example, you may find yourself needing to calculate the Hardy-Weinberg Equillibrium genotype frequencies of a population given the allele frequencies. We can wrap up all the code that you would need to calculate this in a function that we can call upon again and again.

```{r functions}
calc_HWE_geno <- function(p = 0.5){ 
  q <- 1-p
  
  pp <- p^2
  pq <- 2*p*q
  qq <- q^2
  
  return(c(pp, pq, qq))
}

log. <- function(number){ 
  return(log(number))
}

calc_HWE_geno(p = 0.1)

#note that in our UDF we assigned a default value to p (p = 0.5). This means that if we do not specify a value for our argument of p, it will default to using that value.

calc_HWE_geno()
```

### Plots

In addition to mathematical operations, R can help with data visualization. Base R has a few useful plotting functions, but popular packages such as ggplot2 give more customization and control to the user.

```{r Plotting}
hist(gapminder$lifeExp)

boxplot(lifeExp ~ continent, data = gapminder) #box plot for the life expectancies of all years per continent
```

### Setting a random seed

R has many functions that use a random number generator to generate an output. For example, the r____ functions (e.g. rbinom, runif) pull numbers from a probability distribution of your choice. In order to create reproducible analyses, it is often advantageous to be able to reliably obtain the same "random" number after running the same function over again. In order to do so, we can set a seed for the random number generator.

```{r seed}
runif(1,0,1) #runif pulls a number from the uniform distribution with a set of given parameters
runif(1,0,1) #we can see that running runif twice gives you differnt results

set.seed(1234) #setting a seed allows us to obtain reproducible results from functions that use the random number generator
runif(1,0,1)

set.seed(1234)
runif(1,0,1)
```

### Reading and writing data in R

Finally, let us address probably one of the most important points when working with statistics in science: how to get the data you have collected into your R environment. For this part of the lesson, we will be working with the bandersnatch.csv file (created by Katie Long) located here: https://raw.githubusercontent.com/anthonyhung/MSTPsummerstatistics/master/data/bandersnatch.csv. If you would like to have your own copy of this dataset, you can open up a terminal window and run the commands.
```{bash results = 'hide'}
cd ~/Desktop
mkdir data
cd data
wget https://raw.githubusercontent.com/anthonyhung/MSTPsummerstatistics/master/data/bandersnatch.csv
```

Now that we have a copy of the data in a data directory on our desktop, we can load it into R using a relative or absolute directory path and the read.csv function.
```{r read write}
data <- read.csv("~/Desktop/data/bandersnatch.csv")
#let's take a look at the dataset we've just loaded
head(data)
summary(data)

#what is the difference between these two function calls?
head(read.csv("~/Desktop/data/bandersnatch.csv", header = T))
head(read.csv("~/Desktop/data/bandersnatch.csv", header = F))

#let's look at the structure of the data
class(data)
class(data$Color)
class(data$Baseline.Frumiosity)

#let's make some plots with the data
hist(data$Baseline.Frumiosity)
hist(data$Post.Frumiosity)
plot(data$Baseline.Frumiosity, data$Post.Frumiosity)

#we can also write data files and export them using R
data$Size <- rnorm(nrow(data))
write.csv(data, "~/Desktop/data/new_bandersnatch.csv")
```


## Exercises:

1. Write a function called calc_KE that takes as arguments the mass (in kg) and velocity (in m/s) of an object and returns the kinetic energy (in Joules) of an object. Use it to find the KE of a 0.5 kg rock moving at 1.2 m/s. <hide>0.36 Joules</hide>
  
2. Working with the gapminder dataset, find the country with the highest life expectancy in 1962.
<hide>Iceland</hide>

## dplyr

The dplyr package is a powerful tool for working with dataframes (what is a dataframe?) in R. The framework of dplyr relies on the use of "pipes" or the "%>%" operator, to apply multiple useful functions to a dataframe sequentially. Today we will cover these functions:

mutate() adds new variables that are functions of existing variables
select() picks variables based on their names.
filter() picks cases based on their values.
summarise() reduces multiple values down to a single summary.
arrange() changes the ordering of the rows.

However, there are many more useful functions in the package that may come in handy depending on your specific problem!

```{r dplyr intro}
library(dplyr)

gapminder
#mutate
mutate(gapminder, GDP = gdpPercap * pop)
gapminder %>% mutate(GDP = gdpPercap * pop) #pipes

gapminder_with_GDP <- gapminder %>% mutate(GDP = gdpPercap * pop) #assign output to variable
gapminder_with_GDP

#select
gapminder %>% select(c(country, continent, year))
data %>% select(contains("frumiosity"))
gapminder %>%  select(starts_with("c"))
gapminder %>%  select(ends_with("p"))

#filter
gapminder %>% filter(country == "United States")
gapminder %>% filter(year > 1999)
gapminder %>% filter()

#summarise
gapminder %>% summarise(mean(gdpPercap))



#arrange
gapminder %>% arrange(year)
gapminder %>% arrange(desc(year))
gapminder %>% arrange(country, desc(year))


#putting it all together. what if we wanted to know the highest absolute GDP of Zimbabwe between 1982 and 2007? (printing out only 1 value)
gapminder %>% 
  mutate(GDP = gdpPercap * pop) %>% 
  filter(country == "Zimbabwe", between(year, 1982, 2007)) %>% #between!
  arrange(desc(GDP)) %>% 
  select(GDP) %>% 
  top_n(1)
  
# what if we wanted the average GDP of Zimbabwe between 1982 and 2007?
gapminder %>% 
  mutate(GDP = gdpPercap * pop) %>% 
  filter(country == "Zimbabwe", between(year, 1982, 2007)) %>% #between!
  summarise(mean(GDP))

#Bonus: group_by
gapminder %>% 
  group_by(country) %>% 
  summarise(mean(gdpPercap))

gapminder %>% 
  group_by(continent) %>% 
  filter(year == "2007") %>% 
  summarise(n())
```

Exercise: Using dplyr functions, find the Country in Asia that had the largest population in 1982 and tell me it's total GDP that year.

Exercise: The labs data contains information about labs taken for patients over a period of time. Using dplyr, compute the ratio of PaO2/FiO2 for each patient at each time point and output a table of the minimum recorded PaO2/FiO2 ratio for each patient in the data. (not very useful but necessary hint: what is the difference between "PAO2" and "PO2 (ARTERIAL)"?)

```{r exercise labs}
labs <- read.csv("data/labs.csv")
```

